---
title: "Lab3_Initial_Legg"
author: "Legg (Ho Man) Yeung"
date: "July 28, 2017"
output: pdf_document
html_document:
  number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
library(moments)
library(psych)
```


```{r}
unem.data = read.csv("UNRATENSA.csv", header = T)
auto.data = read.csv("TOTALNSA.csv", header = T)
```

# EDA

## Data Overview 

```{r}
str(unem.data)
str(auto.data)
```

```{r}
cbind(head(unem.data),tail(unem.data))
```

```{r}
cbind(head(auto.data),tail(auto.data))
```

```{r}
nrow(unem.data) - nrow(auto.data)
```

- Both datasets are time indexed, accompanied with a key variable of interests. With UNRATENSA, the key variable refers to unemployment rate. With TOTALNSA, the key variable refers to car sale. 

- Both time series presents monthly data. UNRATENSA has 834 observations and starts from 1948-01-01. TOTALNSA has 498 observations  and starts from 1976-01-01. Both ends at 2017-06. UNRATENSA has 28 more years, that is 336 more observations than TOTALNSA.

## Time series Overview

### Time plots and Histograms

```{r}
unem.ts = ts(unem.data$UNRATENSA, frequency = 12, start = c(1948,1))
auto.ts = ts(auto.data$TOTALNSA, frequency = 12, start = c(1976,1))
```

```{r}
ts.plot(unem.ts); title("Unemployment Rate 1948-01 to 2017-06"); 
abline(h = mean(unem.ts), col = "red", lty = "dotted")
#abline(lm(unem.ts~time(unem.ts)), lty = "dotted", col = "blue")
```

(persistent)

```{r}
ts.plot(auto.ts); title("Car Sales 1976-01 to 2017-06")
abline(h = mean(auto.ts), col = "red", lty = "dotted")
#abline(lm(auto.ts~time(auto.ts)), lty = "dotted", col = "blue")
```

(A Sharp Spike in 1980s)
(less persistent than employment rate)

```{r}
hist(unem.data$UNRATENSA, main = "distribution of unemployment rate",
     xlab = "Unemployment Rate")

abline(v = mean(unem.data$UNRATENSA), col = "red", lty = "dotdash"
       , lwd = 2)
abline(v = c((mean(unem.data$UNRATENSA) + sd(unem.data$UNRATENSA)),
             (mean(unem.data$UNRATENSA) + 2*sd(unem.data$UNRATENSA)),
             (mean(unem.data$UNRATENSA) + 3*sd(unem.data$UNRATENSA))), 
       col = "red", lty = "dotted", lwd = 1)
abline(v = c((mean(unem.data$UNRATENSA) - sd(unem.data$UNRATENSA)),
             (mean(unem.data$UNRATENSA) - 2*sd(unem.data$UNRATENSA))), 
       col = "red", lty = "dotted", lwd = 1)

abline(v = median(unem.data$UNRATENSA), col = "green", lty = "dotdash"
       , lwd = 2)

legend("topright", c("Mean","Standard Deviations", "Median"), 
       col = c("red","red","green"), 
       lty = c("dotdash", "dotted","dotdash"))
```

```{r}
hist(auto.data$TOTALNSA, main = "distribution of auto sales",
     xlab = "auto sales - thousands of units")

abline(v = mean(auto.data$TOTALNSA), col = "red", lty = "dotdash"
       , lwd = 2)
abline(v = c((mean(auto.data$TOTALNSA) + sd(auto.data$TOTALNSA)),
             (mean(auto.data$TOTALNSA) + 2*sd(auto.data$TOTALNSA)),
             (mean(auto.data$TOTALNSA) + 3*sd(auto.data$TOTALNSA))), 
       col = "red", lty = "dotted", lwd = 1)
abline(v = c((mean(auto.data$TOTALNSA) - sd(auto.data$TOTALNSA)),
             (mean(auto.data$TOTALNSA) - 2*sd(auto.data$TOTALNSA))), 
       col = "red", lty = "dotted", lwd = 1)

abline(v = median(auto.data$TOTALNSA), col = "green", lty = "dotdash"
       , lwd = 2)

legend("topright", c("Mean","Standard Deviations", "Median"), 
       col = c("red","red","green"), 
       lty = c("dotdash", "dotted","dotdash"))
```


```{r}
summary(unem.data$UNRATENSA)
summary(auto.data$TOTALNSA)
moments::kurtosis(unem.data$UNRATENSA)
```

### Trend Examination with Smoothers and Decomposition

```{r}
# Kernel smoothing
unem.k.smooth.widest = ksmooth(time(unem.ts), 
                             unem.ts, kernel = c("normal"), 
                             bandwidth = 10)

unem.k.smooth.wide = ksmooth(time(unem.ts), 
                             unem.ts, kernel = c("normal"), 
                             bandwidth = 4)

unem.k.smooth.narrow = ksmooth(time(unem.ts), 
                               unem.ts, kernel = c("normal"), 
                               bandwidth = 0.5)

# Make plot
plot(unem.ts, col = "gray", ylab = "Unemployment Rate",
     main = "Unemployment Rate - Kernel Smoothed")
lines(unem.k.smooth.widest$x, unem.k.smooth.widest$y, col = "magenta")
lines(unem.k.smooth.wide$x, unem.k.smooth.wide$y, col = "red")
lines(unem.k.smooth.narrow$x, unem.k.smooth.narrow$y, col = "blue")
abline(lm(unem.ts~time(unem.ts)), lty = "dotted", col = "black")
```

```{r}
# Moving Average Filter
unem.ma.smooth.4year = filter(unem.ts, sides = 1, rep(1/60, 60))
unem.ma.smooth.annual = filter(unem.ts, sides = 1, rep(1/12, 12))
unem.ma.smooth.halfyear = filter(unem.ts, sides = 1, rep(1/6, 6))

# Make plot
plot(unem.ts, col = "gray", ylab = "Unemployment Rate",
     main = "Unemployment Rate - Moving Average Filtered")
lines(unem.ma.smooth.4year, col = "magenta")
lines(unem.ma.smooth.annual, col = "red")
lines(unem.ma.smooth.halfyear, col = "blue")
abline(lm(unem.ts~time(unem.ts)), lty = "dotted", col = "black")
```

```{r}
plot(decompose(unem.ts, type = "additive"))
```

(Decomposition Assumption, Findings of Noise of Inconsistent Variance)

```{r}
# Kernel smoothing
auto.k.smooth.widest = ksmooth(time(auto.ts), 
                             auto.ts, kernel = c("normal"), 
                             bandwidth = 10)

auto.k.smooth.wide = ksmooth(time(auto.ts), 
                             auto.ts, kernel = c("normal"), 
                             bandwidth = 4)

auto.k.smooth.narrow = ksmooth(time(auto.ts), 
                               auto.ts, kernel = c("normal"), 
                               bandwidth = 0.5)

# Make plot
plot(auto.ts, col = "gray", ylab = "Auto Sales thousands of units",
     main = "Car Sales - Kernel Smoothed")
lines(auto.k.smooth.widest$x, auto.k.smooth.widest$y, col = "magenta")
lines(auto.k.smooth.wide$x, auto.k.smooth.wide$y, col = "red")
lines(auto.k.smooth.narrow$x, auto.k.smooth.narrow$y, col = "blue")
abline(lm(auto.ts~time(auto.ts)), lty = "dotted", col = "black")
```

```{r}
# Moving Average Filter
auto.ma.smooth.4year = filter(auto.ts, sides = 1, rep(1/60, 60))
auto.ma.smooth.annual = filter(auto.ts, sides = 1, rep(1/12, 12))
auto.ma.smooth.halfyear = filter(auto.ts, sides = 1, rep(1/6, 6))

# Make plot
plot(auto.ts, col = "gray", ylab = "Car Sales",
     main = "Car Sales - Moving Average Filtered")
lines(auto.ma.smooth.4year, col = "magenta")
lines(auto.ma.smooth.annual, col = "red")
lines(auto.ma.smooth.halfyear, col = "blue")
abline(lm(auto.ts~time(auto.ts)), lty = "dotted", col = "black")
```

```{r}
plot(decompose(auto.ts, type = "additive"))
```
